---
title: "Data Science Capstone Project - Full Analysis"
author: "JJ Sung"
date: "4/20/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Clean workspace
```{r clean}
rm(list=ls())
```

## Packages
```{r packages}
library(rgdal)
library(spatstat)
library(raster)
library(tigris)
library(ggplot2)
library(broom)
library(tidyverse)
library(ggmap)
library(spdep)
library(sp)
library(tidycensus)
library(maditr)
library(spatialreg)
library(gridExtra)
library(rgeos)
library(readxl)
library(devtools)
devtools::install_github("paezha/Spatial-Statistics-Course", subdir = "geog4ga3")
library(geog4ga3)
library(car)
library(MASS)
library(leaps)
```

## US Vaccination Sites
```{r vacc}
#load the dataset
us_vaccSites <- read.csv("Covid19_Vacc_Provider_Locations_US_122921.csv", 
                         header = TRUE)

#descriptive statistics
class(us_vaccSites)
summary(us_vaccSites) #37890 sites

#remove duplicated points
length(us_vaccSites$X)
us_vaccSites <- us_vaccSites[-which(duplicated(us_vaccSites[,c("X","Y")])),]
length(us_vaccSites$X) #1420 locations removed

#clean data so only contiguous US States
length(us_vaccSites$X)
us_vaccSites <- us_vaccSites[!(us_vaccSites$X <= -125),]
length(us_vaccSites$X) #239 removed

#geocoding NA's
length(which(is.na(us_vaccSites$X)))
us_vaccSites_geocode <- us_vaccSites[which(is.na(us_vaccSites$X)),]

#omit NA (no information available)
length(us_vaccSites$X)
us_vaccSites <- us_vaccSites %>% filter(!is.na(us_vaccSites$X))
length(which(is.na(us_vaccSites$X)))
length(us_vaccSites$X) #36230 remaining

#assign initial projection
coordinates(us_vaccSites) <- ~X+Y
proj4string(us_vaccSites)
```

## Brief comparison between old and recent datasets
```{r comparison}
#old1 (April 16)
us_vaccSites_apr <- read.csv("Covid-19_Vaccination_Provider_Locations_in_the_United_States.csv", 
                             header=TRUE)
length(us_vaccSites_apr$X) #21793
us_vaccSites_apr <- us_vaccSites_apr[-which(duplicated(us_vaccSites_apr[,c("X","Y")])),]
us_vaccSites_apr <- us_vaccSites_apr[!(us_vaccSites_apr$X <= -125),]
us_vaccSites_apr <- us_vaccSites_apr %>% filter(!is.na(us_vaccSites_apr$X))
#20889 remaining


#old2 (May 19)
us_vaccSites_may <- read.csv("Covid-19_Vaccination_Provider_Locations_05.19.21.csv", 
                             header=TRUE)
length(us_vaccSites_may$X) #32957
us_vaccSites_may <- us_vaccSites_may[-which(duplicated(us_vaccSites_may[,c("X","Y")])),]
us_vaccSites_may <- us_vaccSites_may[!(us_vaccSites_may$X <= -125),]
us_vaccSites_may <- us_vaccSites_may %>% filter(!is.na(us_vaccSites_may$X))
#31706 remaining


#new (Dec 29)
us_vaccSites_dec <- read.csv("Covid19_Vacc_Provider_Locations_US_122921.csv", 
                         header = TRUE)
length(us_vaccSites_dec$X) #37890
us_vaccSites_dec <- us_vaccSites_dec[-which(duplicated(us_vaccSites_dec[,c("X","Y")])),]
us_vaccSites_dec <- us_vaccSites_dec[!(us_vaccSites_dec$X <= -125),]
us_vaccSites_dec <- us_vaccSites_dec %>% filter(!is.na(us_vaccSites_dec$X))
#36230 remaining

#plot old and new datasets
us_map <- ggmap(get_map(c(left = -125.5, bottom = 24.90, 
                          right = -65.3, top = 49.6), source = "stamen"))
us_map + geom_point(data = us_vaccSites_apr, aes(x = X, y = Y), 
                    size = 0.3, col = 'blue', main = "April 16th data")
us_map + geom_point(data = us_vaccSites_may, aes(x = X, y = Y), 
                    size = 0.3, col = 'blue', main = "May 19th data")
us_map + geom_point(data = us_vaccSites_dec, aes(x = X, y = Y), 
                    size = 0.3, col = 'blue', main = "Dec 29th data")
```

## County borders data
```{r county}
#Importing shapefile
county_shape <- rgdal::readOGR(dsn = "../Capstone/cb_2018_us_county_500k/cb_2018_us_county_500k.shp")
length(which(is.na(county_shape@data))) #0
nrow(county_shape@data) #3233

#Match vaccination & county projections
proj4string(us_vaccSites) <- proj4string(county_shape)

#Create a new column: FIPS (unique code for counties)
county_shape@data$FIPS <- paste(county_shape@data$STATEFP, 
                                county_shape@data$COUNTYFP, sep = "")
```

## Median household income level data
```{r medincome}
#import data
med_income <- read_excel("est19all.xls")

#extract geocode and relevant median income data
med_income <- med_income[-c(1,2,3), c(1,2,23)]

#rename columns
colnames(med_income) <- c("StateFIPS", "CountyFIPS", "med_income")

#remove state/US rows
length(med_income$StateFIPS)
med_income <- med_income[!(med_income$CountyFIPS == "000"),]
length(med_income$StateFIPS)

#Create "FIPS" column
med_income$FIPS <- paste(med_income$StateFIPS, med_income$CountyFIPS, sep = "")
nrow(med_income)
length(which(is.na(med_income$med_income)))

#join to county_shape by FIPS
county_shape@data <- left_join(county_shape@data, med_income)
length(unique(county_shape@data$StateFIPS))

#get rid of counties with no med_income data available
county_shape <- county_shape[-which(is.na(county_shape@data$med_income)),]
length(unique(county_shape@data$STATEFP))
nrow(county_shape@data)
```

## County population by characteristic
```{r popchar}
#load dataset
county_pop_char <- read.csv("cc-est2019-alldata.csv", header = TRUE)
nrow(county_pop_char)

#extract most recent info (07.01.2019)
county_pop_char <- county_pop_char %>% filter(county_pop_char$YEAR == 12)

#format STATE and COUNTY columns
county_pop_char$COUNTY <- as.integer(county_pop_char$COUNTY)
county_pop_char$COUNTY <- sprintf("%03d", county_pop_char$COUNTY)
county_pop_char$STATE <- sprintf("%02d", county_pop_char$STATE)

#create FIPS column
county_pop_char$FIPS <- paste(county_pop_char$STATE, 
                              county_pop_char$COUNTY, sep = "")

#aggregate total population, female, white-only male and female populations
#(group rows by FIPS, city name, and state name)
test <- county_pop_char %>% group_by(CTYNAME, STNAME, FIPS) %>%
        summarise(sum_tot_pop = sum(TOT_POP),
              sum_tot_female = sum(TOT_FEMALE),
              sum_tot_wam = sum(WA_MALE),
              sum_tot_waf = sum(WA_FEMALE))

#create percentage variables (white population, POC population)
test$perc_poc <- 1 - ((test$sum_tot_waf + test$sum_tot_wam)/test$sum_tot_pop)
test$perc_female <- test$sum_tot_female/test$sum_tot_pop

#join to county_shape data
county_shape@data <- left_join(county_shape@data, test)
length(unique(county_shape@data$STATEFP))

county_shape$med_income <- as.numeric(county_shape$med_income)
```

## Aggregating vaccination data by county
```{r agg}
#overlay vacc & county data
over_data <- over(us_vaccSites, county_shape)
length(which(is.na(over_data$GEOID)))

length(us_vaccSites)
length(unique(county_shape@data$StateFIPS))

#aggregate vaccination site by county using GEOID
agg_data <- plyr::count(over_data, c('GEOID'))
agg_data$GEOID <- as.factor(agg_data$GEOID)

#rename columns
colnames(agg_data) <- c("GEOID", "num_vacc_sites")

#add the aggregated column to county_shape@data
county_shape@data <- left_join(county_shape@data, agg_data, 
                               by = (GEOID = "GEOID"))
length(which(is.na(county_shape@data$num_vacc_sites)))

#counties with the biggest number of vaccination sites
county_shape@data$CTYNAME[which(county_shape@data$num_vacc_sites == 
                                        max(na.rm=T, 
                                            county_shape@data$num_vacc_sites))]

#counties with the smallest number of vaccination sites
county_shape@data$CTYNAME[which(county_shape@data$num_vacc_sites == 
                                        min(na.rm=T, 
                                            county_shape@data$num_vacc_sites))]
nrow(county_shape@data)

#tidy up the data and join to county_shape@data
county_fortify <- broom::tidy(county_shape)
county_shape$id <- row.names(county_shape)
county_fortify <- left_join(county_fortify, county_shape@data)
```

## Plot
```{r plot}
#histograms to see the distribution of dependent + independent variables
hist(county_shape@data$num_vacc_sites, 
     main = "Histogram of Number of Vaccination Sites",
     xlab = "Number of Vaccination Sites")
hist(county_shape@data$perc_female,
     main = "Histogram of Female Percentage",
     xlab = "Female Percentage")
hist(county_shape@data$perc_poc,
     main = "Histogram of BIPOC Percentage",
     xlab = "BIPOC Percentage")
hist(county_shape@data$med_income,
     main = "Histogram of Median Household Income",
     xlab = "Median Household Income")

#histogram of logged independent and dependent variables
hist(log(county_shape@data$num_vacc_sites), 
     main = "Histogram of Number of Vaccination Sites (log)",
     xlab = "Number of Vaccination Sites (log)")
hist(log(county_shape@data$perc_female),
     main = "Histogram of Female Percentage (log)",
     xlab = "Female Percentage (log)")
hist(log(county_shape@data$perc_poc),
     main = "Histogram of BIPOC Percentage (log)",
     xlab = "BIPOC Percentage (log)")
hist(log(county_shape@data$med_income),
     main = "Histogram of Median Household Income (log)",
     xlab = "Median Household Income ")

#US map
us_map <- ggmap(get_map(c(left = -125.5, bottom = 24.90, 
                          right = -65.3, top = 49.6), 
                        source = "stamen"))

#choropleth of vaccination sites by county
us_map + geom_polygon(data = county_fortify, 
                      aes(x = long, y = lat, group = group, 
                          fill = log(num_vacc_sites)), 
                      col = 'grey', size = 0.1) + 
        labs(fill = "Number of Vaccination Sites (log)") + 
        scale_fill_distiller(palette = "Spectral")

#choropleth of median household income by county
us_map + geom_polygon(data = county_fortify, 
                      aes(x = long, y = lat, group = group, 
                          fill = log(med_income)), col = 'grey', size = 0.1) + 
        labs(fill = "Median Household Income (log)") + 
        scale_fill_distiller(palette = "Spectral")

#choropleth of total population by county
us_map + geom_polygon(data = county_fortify, 
                      aes(x = long, y = lat, group = group, 
                          fill = log(sum_tot_pop)), col = 'grey', size = 0.1) + 
        labs(fill = "Total Population (log)") + 
        scale_fill_distiller(palette = "Spectral")

#choropleth of total female population percentage by county
us_map + geom_polygon(data = county_fortify, 
                      aes(x = long, y = lat, group = group, fill = perc_female), 
                      col = 'grey', size = 0.1) + 
        labs(fill = "Female Percentage") + 
        scale_fill_distiller(palette = "Spectral")
us_map + geom_polygon(data = county_fortify, 
                      aes(x = long, y = lat, group = group, 
                          fill = log(perc_female)), 
                      col = 'grey', size = 0.1) + 
        labs(fill = "Female Percentage (log)") + 
        scale_fill_distiller(palette = "Spectral")

#choropleth of total POC population percentage by county
us_map + geom_polygon(data = county_fortify, 
                      aes(x = long, y = lat, group = group, 
                          fill = log(perc_poc)), col = 'grey', size = 0.1) + 
        labs(fill = "BIPOC Percentage (log)") + 
        scale_fill_distiller(palette = "Spectral")
```

## Areal models
## Initial full models
```{r linreg}
#remove counties with no vaccination sites (NA's) for analysis
county_shape_ref <- county_shape@data[-which
                                      (is.na
                                        (county_shape@data$num_vacc_sites)),]

attach(county_shape_ref)

#first-order Multiple Linear Regression (MLR)
lin_reg <- lm(log(num_vacc_sites)~med_income+sum_tot_pop+perc_poc+perc_female, 
              data=county_shape_ref)
summary(lin_reg)

#Check multicollinearity
vif(lin_reg) #all values smaller than 2, no multicollinearity

#Full interaction model
int_var <- subset(county_shape_ref, select = c(num_vacc_sites,med_income,
                                                sum_tot_pop,perc_poc,
                                                perc_female))
full_int <- lm(log(num_vacc_sites)~.:., data = int_var)
summary(full_int)
```

## Variable Selection: First-order model
```{r}
#Stepwise selection with AIC criterion
aic.fo.mod <- stepAIC(lin_reg, direction = "both", trace = 0, k = 2)
summary(aic.fo.mod)

#Stepwise selection with BIC criterion
bic.fo.mod <- stepAIC(lin_reg, direction = "both", trace = 0, k = log(50))
summary(bic.fo.mod)

#All-subset selection
detach(county_shape_ref)
attach(int_var)
x.mod <- cbind(int_var$med_income,
               int_var$sum_tot_pop,
               int_var$perc_poc,
               int_var$perc_female)
y.mod <- int_var$num_vacc_sites
leap.out <- leaps(x.mod, y.mod, method = c("r2"), nbest = 1)
leap.out$which[leap.out$size==3] #R-squared

leap.out <- leaps(x.mod, y.mod, method = c("Cp"), nbest = 1)
leap.out$which[leap.out$size==3] #Mallow's Cp
detach(int_var)
```

## Variable Selection: Interaction model
```{r}
#Stepwise selection with AIC criterion
aic.int.mod <- stepAIC(full_int, direction = "both", trace = 0, k = 2)
summary(aic.int.mod)

#Stepwise selection with BIC criterion
bic.int.mod <- stepAIC(full_int, direction = "both", trace = 0, k = log(50))
summary(bic.int.mod)
```

## Cross validation
```{r}
library(caret)
set.seed(123)

#10-fold CV
ctrl <- trainControl(method = "cv",number=10)
cv.fo.model1 <- train(num_vacc_sites~med_income+sum_tot_pop
                      +perc_poc+perc_female,data = int_var, 
                      method = "lm", trControl = ctrl)
cv.fo.model2 <- train(num_vacc_sites~sum_tot_pop+perc_female, 
                      data = int_var, method = "lm", trControl = ctrl)
cv.bic.int.model<- train(num_vacc_sites ~ med_income+sum_tot_pop
                         +perc_poc+perc_female+med_income:sum_tot_pop
                         +med_income:perc_poc+med_income:perc_female
                         +sum_tot_pop:perc_poc +sum_tot_pop:perc_female, 
                         data = int_var,method="lm",trControl=ctrl)

print(cv.fo.model1)
print(cv.fo.model2)
print(cv.bic.int.model)
```

## Spatial error model
```{r serror}
#create neighborhood matrix
set.seed(3)
#copy county_shape data for neighborhood matrix
county_shape_nb <- county_shape
length(county_shape_nb)
length(which(is.na(county_shape_nb@data)))

us_nb <- poly2nb(county_shape_nb)
W_list <- nb2listw(us_nb, style = "W", zero.policy = TRUE)
summary(us_nb)

length(lin_reg$residuals)
length(us_nb)

#Moran's I
moran.mc(lin_reg$residuals, listw = W_list, zero.policy = TRUE, nsim = 5000)

#SAR model
vacc_sar <- errorsarlm(log(num_vacc_sites) ~ med_income + sum_tot_pop + 
                         perc_female + perc_poc, 
                       data = county_shape@data, 
                       listw = W_list, zero.policy = TRUE)
summary(vacc_sar) #AIC: 6463.9, (AIC for lm: 6836.1)
```

## Spatial Lag Model
```{r slag}
vacc_lag <- lagsarlm(log(num_vacc_sites) ~ med_income + sum_tot_pop + 
                       perc_female + perc_poc, data = county_shape@data, 
                     listw = W_list, zero.policy = TRUE)
summary(vacc_lag) #AIC: 6487.1
```

## CAR
```{r car}
vacc_car <- spautolm(log(num_vacc_sites) ~ med_income + sum_tot_pop + 
                       perc_female + perc_poc, data = county_shape@data, 
                     family = "CAR", listw = W_list, zero.policy = TRUE)
summary(vacc_car) #AIC: 6411.9
```

## Point Process Model
```{r gnf}
set.seed(3)
range(us_vaccSites$X)
range(us_vaccSites$Y)

#California
vacc_ca <- us_vaccSites[which(us_vaccSites$State == "CA"),]
vacc_ca2 <- spTransform(vacc_ca, CRS("+init=epsg:2225"))

#set window
county_ca <- county_shape[which(county_shape@data$StateFIPS == "06"),]
ca_poly <- as(county_ca, "SpatialPolygons") #convert to SpatialPolygons
ca_poly <- spTransform(ca_poly, CRS("+init=epsg:2225"))
ca_one <- st_transform(st_union(st_as_sf(ca_poly)), 
                       scrs = CRS("+init=epsg:2225"))
ca_win <- as.owin(as_Spatial(ca_one))

ca_ppp <- as.ppp(vacc_ca2@coords, W = ca_win)
#plot point process model
plot(ca_ppp, main = "CA vaccination sites")

#G-function
g_env_ca <- envelope(ca_ppp, fun = Gest, nrank = 2, nsim = 99)
plot(g_env_ca, main = "CA vaccination sites - G function", 
     xlim = c(0,max(g_env_ca$r)))

#F-function
f_env_ca <- envelope(ca_ppp, fun = Fest, nrank = 2, nsim = 99)
plot(f_env_ca, main = "CA vaccination sites - F function", 
     xlim = c(0,max(f_env_ca$r)))

#New York
vacc_ny <- us_vaccSites[which(us_vaccSites$State == "NY"),]
vacc_ny2 <- spTransform(vacc_ny, CRS("+init=epsg:2261"))

#set window
county_ny <- county_shape[which(county_shape@data$StateFIPS == "36"),]
ny_poly <- as(county_ny, "SpatialPolygons") 
ny_poly <- spTransform(ny_poly, CRS("+init=epsg:2261"))
ny_one <- st_transform(st_union(st_as_sf(ny_poly)), crs = CRS("+init=epsg:2261"))
ny_win <- as.owin(as_Spatial(ny_one))

ny_ppp <- as.ppp(vacc_ny2@coords, W = ny_win)
#plot point process model
plot(ny_ppp, main = "NY vaccination sites")

#G-function
g_env_ny <- envelope(ny_ppp, fun = Gest, nrank = 2, nsim = 99)
plot(g_env_ny, main = "NY vaccination sites - G function", 
     xlim = c(0,max(g_env_ny$r)))

#F-function
f_env_ny <- envelope(ny_ppp, fun = Fest, nrank = 2, nsim = 99)
plot(f_env_ny, main = "NY vaccination sites - F function", 
     xlim = c(0,max(f_env_ny$r)))

#Rural areas
#Alabama
vacc_al <- us_vaccSites[which(us_vaccSites$State == "AL"),]
vacc_al2 <- spTransform(vacc_al, CRS("+init=epsg:2759"))

#set window
county_al <- county_shape[which(county_shape@data$StateFIPS == "01"),]
al_poly <- as(county_al, "SpatialPolygons")
al_poly <- spTransform(al_poly, CRS("+init=epsg:2759"))
al_one <- st_transform(st_union(st_as_sf(al_poly)), crs = CRS("+init=epsg:2759"))
al_win <- as.owin(as_Spatial(al_one))

al_ppp <- as.ppp(vacc_al2@coords, W = al_win)

#plot point process model
plot(al_ppp, main = "AL vaccination sites")

#G-function
g_env_al <- envelope(al_ppp, fun = Gest, nrank = 2, nsim = 99)
plot(g_env_al, main = "AL vaccination sites - G function", 
     xlim = c(0,max(g_env_al$r)))

#F-function
f_env_al <- envelope(al_ppp, fun = Fest, nrank = 2, nsim = 99)
plot(f_env_ca, main = "AL vaccination sites - F function", 
     xlim = c(0,max(f_env_al$r)))

#Kentucky
vacc_ky <- us_vaccSites[which(us_vaccSites$State == "KY"),]
vacc_ky2 <- spTransform(vacc_ky, CRS("+init=epsg:3088"))

#set window
county_ky <- county_shape[which(county_shape@data$StateFIPS == "21"),]
ky_poly <- as(county_ky, "SpatialPolygons")
ky_poly <- spTransform(ky_poly, CRS("+init=epsg:3088"))
ky_one <- st_transform(st_union(st_as_sf(ky_poly)), crs = CRS("+init=epsg:3088"))
ky_win <- as.owin(as_Spatial(ky_one))

ky_ppp <- as.ppp(vacc_ky2@coords, W = ky_win)

#plot point process model
plot(ky_ppp, main = "KY vaccination sites - F function")

#G-function
g_env_ky <- envelope(ky_ppp, fun = Gest, nrank = 2, nsim = 99)
plot(g_env_ky, main = "KY vaccination sites - G function", 
     xlim = c(0,max(g_env_ky$r)))

#F-function
f_env_ky <- envelope(ky_ppp, fun = Fest, nrank = 2, nsim = 99)
plot(f_env_ky, main = "KY vaccination sites - F function", 
     xlim = c(0,max(f_env_ky$r)))
```

